---
title: "Case study: reproducing a Metabolite Merging Strategy with MetMashR"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Case study: reproducing a Metabolite Merging Strategy with MetMashR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center"
)

.DT <- function(x) {
    DT::datatable(
        x,
        options = list(
            scrollX = TRUE,
            pageLength = 10,
            dom = "ftip"
        ),
        rownames = FALSE
    )
}
```

```{r setup, message = FALSE}
library(struct)
library(MetMashR)
library(ggplot2)
```

# Introduction

Villalba et al. (2023) describe a "Metabolites Merging Strategy" (MMS): a
three-step procedure for harmonising metabolite annotations reported by
different studies, databases and publications into one consistent,
cross-referenced dataset. The steps are:

1. **Translation and merging** - convert whatever identifier each source
   reports (a name, an InChIKey, or some other identifier) to a common
   InChIKey, and merge records that share one.
2. **Attribute retrieval** - for each InChIKey, retrieve descriptive
   attributes (name, molecular weight/formula) and cross-reference
   identifiers (PubChem CID, InChI, SMILES, ChEBI, HMDB, KEGG, LipidMaps,
   DrugBank, CAS) and chemical ontology (kingdom/superclass/class).
3. **Manual curation** - collapse conjugated acid/base forms of the same
   compound, retrieve missing attributes, and resolve duplicate/conflicting
   synonyms.

Their own worked example harmonises urinary asthma metabolites reported
across three Metabolomics Workbench studies (ST001039, ST001048, ST001317),
the Human Metabolome Database, and six literature articles.

This vignette reproduces the MWB portion of that case study as a **single
MetMashR model sequence** - one workflow object built by chaining existing
(and, where a genuine gap existed, newly added) MetMashR steps with `+` -
and checks the result against the paper's own published table.

## A note on what "translation" means here

Villalba et al.'s Step 1 used the PubChem Identifier Exchange Service for
names, and the Chemical Translation Service (CTS) for other identifier
types. **CTS has since been shut down** - every endpoint of the original
service now returns a migration notice, and its replacement, CTS-Lite, does
not support name-to-identifier translation at all (it only matches an
identifier you already have against its own database). This vignette uses
[Metabolomics Workbench's RefMet
service](https://www.metabolomicsworkbench.org/databases/refmet/) as a
substitute: it plays a similar synonym-resolution role, is still live, and
in practice its curated picks often line up with the biologically-relevant
form of a compound (the correct stereoisomer or protonation state) more
often than a plain PubChem name search does - see the comparison in
Section \@ref(step1) below.

# Step 0: import the source data

Each study's reported metabolite list is pulled directly from Metabolomics
Workbench with `mwb_study_source()`, and the three tables are stacked with
`combine_sources()`.

Only the `metabolite_name` column - the name as originally reported by each
study - is carried forward into the workflow below. Metabolomics Workbench
does attach a `refmet_name` to most rows automatically at deposit time, and
occasionally a `pubchem_id`, but using those would trivialise the exercise:
the scenario MetMashR's mashing steps are built for is annotation software
output that has names only, with no standardised identifiers attached. Any
pre-existing identifiers are kept aside purely so the final comparison in
Section \@ref(comparison) has a well-defined published answer to check
against - never fed into the translation steps themselves.

```{r step0, eval = FALSE}
study_ids <- c("ST001039", "ST001048", "ST001317")
sources <- lapply(
    study_ids,
    function(id) read_source(mwb_study_source(source = id, tag = id))
)

combine_step <- combine_sources(
    source_list = list(),
    keep_cols = "metabolite_name",
    source_col = "source"
)
combined <- predicted(model_apply(combine_step, sources))
```

```{r step0-run, echo = FALSE}
study_ids <- c("ST001039", "ST001048", "ST001317")
```

# Building the model sequence {#step1}

The remaining steps are chained into one `model_seq` object with `+`. To
avoid overburdening the public APIs involved every time this vignette is
built, each step that queries an external service is given an `rds_cache`
pointing at a result set cached with the package (`inst/cached/`); in your
own workflow you would typically omit `cache` (or point it at your own
file) so that your queries populate it as you go. Beyond speed, caching
also leaves a persistent local record of what a service actually returned
- useful given that services used here can be discontinued or changed with
no notice, as happened to CTS (Section \@ref(step1)).

```{r cache-setup, eval = FALSE}
cached <- function(name) {
    rds_cache(source = system.file(
        "cached", package = "MetMashR",
        paste0("mms_case_study_", name, ".rds")
    ))
}
```

## Step 0.5: normalising synonyms

A handful of normalisation rules, developed by inspecting real name-match
failures, are applied before translation:

- extract a name from its trailing parenthetical gloss when one is present
  and preceded by whitespace (e.g. `"MEP (Mono-ethyl phthalate)"` ->
  `"Mono-ethyl phthalate"`) - guarding on the preceding space keeps lipid
  shorthand like `"LysoPC(18:0)"` (no space before the paren) untouched;
- strip racemic/stereo markers MetMashR's built-in `racemic_dictionary`
  doesn't already cover (`"D/L "`, `"DL-"` and their mirror forms);
- expand the `Lyso-` glycerophospholipid class prefix to LIPID MAPS'
  shorthand (`"LysoPC"` -> `"LPC"`), which resolves cleanly where the full
  word often doesn't.

```{r step0.5, eval = FALSE}
abbrev_dictionary <- list(
    list(
        pattern = "\\s\\([^()]+\\)\\s*$",
        replace = function(inp) {
            m <- regmatches(
                inp$x,
                regexec("\\s\\(([^()]+)\\)\\s*$", inp$x)
            )
            if (length(m[[1]]) > 1) trimws(m[[1]][2]) else inp$x
        }
    )
)
dl_dictionary <- list(
    list(pattern = "^D/L[ -]?", replace = ""),
    list(pattern = "^L/D[ -]?", replace = ""),
    list(pattern = "^DL-?", replace = ""),
    list(pattern = "^LD-?", replace = "")
)
lyso_dictionary <- list(
    list(pattern = "^Lyso-?PC", replace = "LPC", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PE", replace = "LPE", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PI", replace = "LPI", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PG", replace = "LPG", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PS", replace = "LPS", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PA", replace = "LPA", perl = TRUE, ignore.case = TRUE)
)

normalise_step <- normalise_strings(
    search_column = "metabolite_name",
    output_column = "name_normalised",
    dictionary = c(abbrev_dictionary, racemic_dictionary, dl_dictionary, lyso_dictionary)
)
```

## Step 1: translation and merging

Two independent translation routes are tried for every (normalised) name:
[`pubchem_id_exchange()`], which drives PubChem's own batch PUG XML ID
Exchange service directly, and [`mwb_refmet_lookup()`], the RefMet-based
substitute for CTS described above. [`prioritise_columns()`] then picks a
winner per row, **preferring RefMet's pick when one exists**: across the
name-resolution problems encountered developing this vignette, RefMet
consistently resolved to the same biologically-relevant
stereoisomer/protonation state as the paper's own curated table, where a
plain PubChem name search more often returned a generic, stereochemistry-
unspecified, or wrongly-charged form of the same compound.

Each phase below is built as a plain `list()` of steps rather than
pre-chaining it into its own `model_seq` with `+`: `struct` defines `+` for
`model`/`model` and `model`/`model_seq` combinations, but not for combining
two already-built `model_seq` objects together, which is what naming each
phase and joining the phases with `+` at the end would require. Keeping
each phase a list and flattening with `Reduce("+", ...)` once, at the very
end (Section \@ref(the-complete-sequence)), sidesteps that gap while still
keeping the workflow readable in named, logical chunks.

```{r step1, eval = FALSE}
translate_step <- list(
    pubchem_id_exchange(
        query_column = "name_normalised",
        input_type = "synonyms",
        output_type = "inchikey",
        cache = cached("pubchem_id_exchange")
    ),
    mwb_refmet_lookup(
        query_column = "name_normalised",
        cache = cached("refmet"),
        delay = 0.5
    ),
    # flag cases where the two translation routes agree on connectivity
    # (same first-14-character InChIKey block) but disagree on
    # stereochemistry (differ within the first 26 characters) - a signal
    # that automated resolution is unreliable here and the pick genuinely
    # needs human review (MMS Step 3c). See Section \@ref(stereoisomers)
    # for how well this actually catches the mismatches found by checking
    # against the paper's published result.
    compute_column(
        input_columns = c("inchi_key_refmet", "inchikey_pubchem_id_exchange"),
        output_column = "stereo_ambiguous",
        fcn = function(x) {
            a <- x[[1]]
            b <- x[[2]]
            ambiguous <- substr(a, 1, 14) == substr(b, 1, 14) &
                substr(a, 1, 26) != substr(b, 1, 26)
            ambiguous[is.na(ambiguous)] <- FALSE
            ambiguous
        }
    ),
    prioritise_columns(
        column_names = c("inchi_key_refmet", "inchikey_pubchem_id_exchange"),
        output_name = "InChIKey",
        source_name = "InChIKey_source",
        source_tags = c("refmet", "pubchem")
    ),
    filter_na(column_name = "InChIKey", mode = "exclude"),
    combine_records(
        group_by = "InChIKey",
        default_fcn = fuse_unique(separator = " || "),
        fcns = list(
            metabolite_name = fuse_unique(separator = " || "),
            source = fuse_unique(separator = " || "),
            stereo_ambiguous = function(x) any(x, na.rm = TRUE)
        )
    )
)
```

Names that are already an InChIKey, or reported as some other identifier
type, pass straight through unchanged: only rows without a value in
`InChIKey` are actually queried by the translation steps above, since
`pubchem_id_exchange()`/`mwb_refmet_lookup()` only touch their configured
`query_column`. None of the three MWB studies used here happen to report
anything other than a plain name, so this vignette doesn't exercise that
path directly, but no special-casing is needed for it either.

## Step 2: attribute retrieval

Each translated InChIKey is enriched with: name, molecular weight/formula,
InChI and SMILES from PubChem (also the source for molecular
weight/formula, since CTS - the paper's original source for these - is no
longer available); chemical ontology from ClassyFire (needs a structure,
so this step runs after the PubChem lookup above supplies one); HMDB, ChEBI
and LipidMaps identifiers from Metabolomics Workbench's own compound
database (`mwb_compound_lookup()`, a role CTS also used to fill); and a
KEGG identifier by chaining through the PubChem CID.

ClassyFire's ontology lookup uses [`classyfire_batch_lookup()`] rather than
[`classyfire_lookup()`]: the latter is a synchronous, one-compound-per-
request call, and ClassyFire rate-limits it aggressively enough to make
several hundred compounds impractical. `classyfire_batch_lookup()` instead
uses [ClassyFire's own batch submission
API](https://bitbucket.org/wishartlab/classyfire_api/src/master/) - one (or
a few, for very large inputs) request covers the whole batch.

DrugBank ID and CAS number, part of the paper's original attribute set, are
omitted: DrugBank has no accessible public REST API for this kind of
lookup, and while PubChem's `xrefs` endpoint can return a CAS Registry
Number for many compounds, no MetMashR object currently wraps it.

```{r step2, eval = FALSE}
attribute_step <- list(
    pubchem_property_lookup(
        query_column = "InChIKey",
        search_by = "inchikey",
        property = c(
            "IUPACName", "MolecularWeight", "MolecularFormula",
            "InChI", "CanonicalSMILES", "Charge"
        ),
        cache = cached("pubchem_property"),
        delay = 0.4
    ),
    classyfire_batch_lookup(
        query_column = "ConnectivitySMILES_pubchem",
        output_items = c("kingdom", "superclass", "class"),
        output_fields = "name",
        cache = cached("classyfire"),
        delay = 3
    ),
    mwb_compound_lookup(
        input_item = "inchi_key",
        query_column = "InChIKey",
        output_item = c("hmdb_id", "chebi_id", "lm_id", "pubchem_cid"),
        cache = cached("mwb_compound"),
        delay = 0.5
    ),
    kegg_lookup(
        get = "compound", from = "pubchem",
        query_column = "CID_pubchem"
    )
)
```

## Step 3: manual curation

Villalba et al. describe collapsing conjugated acid/base forms using "the
first 14 characters" of the InChIKey (the connectivity/skeleton hash only).
Tested against real data, that rule also merges unrelated stereoisomers
that happen to share a skeleton (D-glucuronic acid and its epimer
galacturonate, both neutral, is one real example). An InChIKey's *final*
single character is specifically a protonation flag (`N` = neutral, other
letters = a specific charge state) - verified here against live PubChem
`Charge` values - so grouping on everything **except** that last character,
then keeping whichever group member has the smallest `|Charge|`, collapses
true conjugate pairs without also merging genuine stereoisomers.

```{r step3, eval = FALSE}
attr_cols <- c(
    "IUPACName_pubchem", "MolecularWeight_pubchem", "MolecularFormula_pubchem",
    "InChI_pubchem", "ConnectivitySMILES_pubchem", "CID_pubchem",
    "hmdb_id_mwb", "chebi_id_mwb", "lm_id_mwb", "pubchem_cid_mwb",
    "compound_kegg", "kingdom.name_cfb", "superclass.name_cfb", "class.name_cfb",
    # keep the winning row's own flag rather than the default fuse_unique(),
    # which would coerce this logical column to a "TRUE"/"FALSE" string
    "stereo_ambiguous"
)
select_min_fcns <- setNames(
    lapply(
        attr_cols,
        function(x) select_min(min_col = "Charge_pubchem", use_abs = TRUE)
    ),
    attr_cols
)

curation_step <- list(
    compute_column(
        input_columns = "InChIKey",
        output_column = "inchikey_skeleton",
        # compute_column passes fcn a 1-column data.frame, not a bare vector
        fcn = function(x) substr(x[[1]], 1, 26)
    ),
    combine_records(
        group_by = "inchikey_skeleton",
        default_fcn = fuse_unique(separator = " || "),
        fcns = select_min_fcns
    ),
    unique_records()
)
```

Missing attributes (a translated InChIKey that some downstream lookup
couldn't find a match for) surface directly as `NA` in the relevant column
- inspect these with `filter_na()` in your own workflow to decide whether
manual review or a further fallback (e.g. `opsin_lookup()`) is warranted.

## The complete sequence

Written out in full, the entire workflow above is one continuous chain of
steps joined with `+` - a single `model_seq`, built and run once against
`combined`:

```{r full-workflow-display, eval = FALSE}
mms_workflow <-
    normalise_strings(
        search_column = "metabolite_name",
        output_column = "name_normalised",
        dictionary = c(abbrev_dictionary, racemic_dictionary, dl_dictionary, lyso_dictionary)
    ) +
    pubchem_id_exchange(
        query_column = "name_normalised",
        input_type = "synonyms",
        output_type = "inchikey",
        cache = cached("pubchem_id_exchange")
    ) +
    mwb_refmet_lookup(
        query_column = "name_normalised",
        cache = cached("refmet"),
        delay = 0.5
    ) +
    compute_column(
        input_columns = c("inchi_key_refmet", "inchikey_pubchem_id_exchange"),
        output_column = "stereo_ambiguous",
        fcn = function(x) {
            a <- x[[1]]
            b <- x[[2]]
            ambiguous <- substr(a, 1, 14) == substr(b, 1, 14) &
                substr(a, 1, 26) != substr(b, 1, 26)
            ambiguous[is.na(ambiguous)] <- FALSE
            ambiguous
        }
    ) +
    prioritise_columns(
        column_names = c("inchi_key_refmet", "inchikey_pubchem_id_exchange"),
        output_name = "InChIKey",
        source_name = "InChIKey_source",
        source_tags = c("refmet", "pubchem")
    ) +
    filter_na(column_name = "InChIKey", mode = "exclude") +
    combine_records(
        group_by = "InChIKey",
        default_fcn = fuse_unique(separator = " || "),
        fcns = list(
            metabolite_name = fuse_unique(separator = " || "),
            source = fuse_unique(separator = " || "),
            stereo_ambiguous = function(x) any(x, na.rm = TRUE)
        )
    ) +
    pubchem_property_lookup(
        query_column = "InChIKey",
        search_by = "inchikey",
        property = c(
            "IUPACName", "MolecularWeight", "MolecularFormula",
            "InChI", "CanonicalSMILES", "Charge"
        ),
        cache = cached("pubchem_property"),
        delay = 0.4
    ) +
    classyfire_batch_lookup(
        query_column = "ConnectivitySMILES_pubchem",
        output_items = c("kingdom", "superclass", "class"),
        output_fields = "name",
        cache = cached("classyfire"),
        delay = 3
    ) +
    mwb_compound_lookup(
        input_item = "inchi_key",
        query_column = "InChIKey",
        output_item = c("hmdb_id", "chebi_id", "lm_id", "pubchem_cid"),
        cache = cached("mwb_compound"),
        delay = 0.5
    ) +
    kegg_lookup(
        get = "compound", from = "pubchem",
        query_column = "CID_pubchem"
    ) +
    compute_column(
        input_columns = "InChIKey",
        output_column = "inchikey_skeleton",
        fcn = function(x) substr(x[[1]], 1, 26)
    ) +
    combine_records(
        group_by = "inchikey_skeleton",
        default_fcn = fuse_unique(separator = " || "),
        fcns = select_min_fcns
    ) +
    unique_records()

result <- predicted(model_apply(mms_workflow, combined))
```

# What the harmonised table looks like

`result$data` is a single flat table: one row per unique compound, with
every attribute from Steps 1-3 as a column - the InChIKey and its
provenance, PubChem/RefMet identifiers, ClassyFire ontology, and
cross-references from MWB, ChEBI, HMDB, LipidMaps and KEGG. A cleaned-up
subset of the columns is shown below (the full table also carries e.g. the
raw ClassyFire class *descriptions*, InChI strings and SMILES, omitted here
just for readability):

```{r result-table, eval = FALSE}
preview_cols <- c(
    "metabolite_name", "InChIKey", "IUPACName_pubchem",
    "MolecularFormula_pubchem", "MolecularWeight_pubchem",
    "hmdb_id_mwb", "chebi_id_mwb", "pubchem_cid_mwb", "compound_kegg", "lm_id_mwb",
    "kingdom.name_cfb", "superclass.name_cfb", "class.name_cfb",
    "stereo_ambiguous", "source"
)
.DT(result$data[, preview_cols])
```

```{r result-table-cached, echo = FALSE}
.DT(readRDS(system.file("cached", "mms_case_study_result_preview.rds", package = "MetMashR")))
```

# Comparison against the published result {#comparison}

Villalba et al.'s own final table is published as supplementary material on
Zenodo (10.5281/zenodo.8226097). It's imported here with
`BiocFileCache_database()` so it isn't re-downloaded on every build, and
restricted to compounds attributed to these three MWB studies (the paper's
HMDB and literature sources are out of scope for this vignette, which only
implements Step 0 for Metabolomics Workbench).

```{r comparison, eval = FALSE}
zenodo_url <- paste0(
    "https://zenodo.org/api/records/8226097/files/",
    "MDM_Suppl_vSubmitted.xlsx/content"
)
gold_db <- BiocFileCache_database(
    source = zenodo_url,
    resource_name = "MMS_zenodo_suppl",
    bfc_fun = cache_as_is,
    import_fun = function(path) {
        # the Zenodo API URL has no .xlsx extension; openxlsx needs one
        tmp <- tempfile(fileext = ".xlsx")
        file.copy(path, tmp, overwrite = TRUE)
        on.exit(unlink(tmp))
        openxlsx::read.xlsx(tmp, sheet = "Table S3", colNames = TRUE)
    }
)
gold_full <- read_database(gold_db)
gold_mwb <- gold_full[grepl(paste(study_ids, collapse = "|"), gold_full$Reference), ]
```

```{r venn, eval = FALSE}
ours <- annotation_source(
    name = "MetMashR", tag = "MetMashR",
    data = data.frame(InChIKey = unique(result$data$InChIKey))
)
gold <- annotation_source(
    name = "MMS (Villalba et al.)", tag = "MMS (Villalba et al.)",
    data = data.frame(InChIKey = unique(gold_mwb$InChIKey))
)

C <- annotation_venn_chart(
    factor_name = "InChIKey", labels = TRUE, legend = TRUE,
    # colour by group with alpha blending in the overlap - annotation_venn_chart()'s
    # own default is a plain white fill; ".group" turns on ggVennDiagram's
    # per-group colour mapping (alpha is fixed internally at 1/n groups)
    fill_colour = ".group"
)
chart_plot(C, ours, gold)
```

```{r venn-image, echo = FALSE, out.width = "70%"}
knitr::include_graphics(
    system.file("cached", "mms_case_study_venn.png", package = "MetMashR")
)
```

# Characterising the mismatch {#stereoisomers}

Rather than leave "55 only ours, 58 only paper" as an unexplained gap,
every mismatched pair was checked at the InChIKey skeleton level: the first
14 characters encode connectivity only, and the first 26 encode
connectivity plus stereochemistry, leaving only the final character (a
protonation flag) unaccounted for.

```{r mismatch-check, eval = FALSE}
our_keys <- unique(result$data$InChIKey)
gold_keys <- unique(gold_mwb$InChIKey)
only_ours <- setdiff(our_keys, gold_keys)
only_gold <- setdiff(gold_keys, our_keys)

skel14 <- function(x) substr(x, 1, 14)
skel26 <- function(x) substr(x, 1, 26)

pairs <- do.call(rbind, lapply(only_ours, function(k) {
    g <- only_gold[skel14(only_gold) == skel14(k) & skel26(only_gold) != skel26(k)]
    if (length(g) == 0) {
        return(NULL)
    }
    data.frame(ours_key = k, gold_key = g)
}))
```

Of the 55 compounds only in MetMashR's result, 36 turned out to share a
skeleton with a paper-only compound but disagree on stereochemistry, 1 is
the `Phosphocholine`/`Phosphorylcholine` cation-vs-neutral case described
above, and 18 have no skeletal relationship to any paper-only compound at
all (likely genuine name-resolution misses).

Breaking the 36 stereo mismatches down further, by which side (if either)
resolved to the InChIKey's "no stereochemistry specified" stereo layer:

| Direction | Count |
|---|---|
| MetMashR resolves to a *specific* stereoisomer, the paper's is unspecified | 24 |
| Both resolve to a *specific* but *different* stereoisomer | 10 |
| MetMashR's is unspecified, the paper's is specific | 2 |

The first row isn't really a defect - MetMashR's output is arguably more
precise than the reference there, not less. The other 12 are the real gap,
and for at least 10 of them the root cause is checkable directly: the
paper's *curated* name carries a stereo descriptor the *originally
reported* name never had (e.g. `Rosmarinate` -> "(S)-rosmarinic acid",
`Lactose` -> "beta-Lactose", `Biopterin` -> "D-Biopterin"). That
stereochemistry was added by a human curator and isn't recoverable from the
bare reported name by any automated lookup.

## How well does `stereo_ambiguous` catch these?

The flag added in Section \@ref(step1) compares MetMashR's own two
translation routes against *each other*, not against the paper, so its
recall is necessarily imperfect: it can only catch disagreement it can see.

```{r flag-check, eval = FALSE}
sum(result$data$stereo_ambiguous, na.rm = TRUE) # rows flagged
sum(pairs$ours_key %in% result$data$InChIKey[result$data$stereo_ambiguous]) # of the 36, how many caught
```

In practice it flags 32 of 335 compounds, catching 27 of the 36 known
mismatches (75%) plus 5 more where the two sources disagreed internally but
the final pick happened to still match the paper anyway (worth review
regardless, since that agreement was coincidental). The 9 it misses are
cases where PubChem ID Exchange and RefMet happen to agree with *each
other* on the wrong stereoisomer - a shared bias that comparing two sources
against one another can't detect, and the honest limit of this approach
without a third, independent source to triangulate against.

## What perfect resolution would look like

To size the ceiling on this specific fix, each of the 36 mismatched
InChIKeys was relabelled to the paper's matching key (simulating what
correct manual or better-informed resolution would produce) and the
overlap recomputed:

```{r resolved-hypothetical, eval = FALSE}
resolved_keys <- our_keys
m <- match(resolved_keys, pairs$ours_key)
hit <- !is.na(m)
resolved_keys[hit] <- pairs$gold_key[m[hit]]
resolved_keys <- unique(resolved_keys)

length(intersect(resolved_keys, gold_keys)) # shared, after resolving
```

```{r venn-resolved-image, echo = FALSE, out.width = "70%"}
knitr::include_graphics(
    system.file("cached", "mms_case_study_venn_resolved.png", package = "MetMashR")
)
```

Shared compounds rise from 280 (83.6%/82.8%) to **316 (93.5%/93.5%)** -
resolving the stereochemistry gap alone would close most of the remaining
distance to the published result, leaving only the 19/22 split from
genuinely unmatched compounds and (on the paper's side) sources this
vignette doesn't cover.

# Summary

Reproducing MMS as one MetMashR model sequence required three genuinely new
objects (`mwb_study_source()` for Step 0, `mwb_refmet_lookup()` as a
substitute for the now-defunct CTS, `classyfire_batch_lookup()` to make
Step 2's ontology retrieval tractable at this scale) and one ported from an
internal sister package (`pubchem_id_exchange()`, a closer match to the
paper's own "PubChem Identifier Exchange Service" than the synchronous
per-row lookups MetMashR already had). Everything else - normalisation,
attribute retrieval, conjugate-form collapsing, deduplication, flagging
disagreement for review - was already available and needed no new code,
only careful chaining and, in a couple of places, a column-name fix or a
widened retry.

The value of building this as one executable, versioned workflow object,
rather than a paraphrased description of one, showed up directly during
development: it surfaced a specific ambiguity in the original method's own
description of conjugate-form collapsing (the literal "first 14
characters" rule over-merges real stereoisomers), a case where the exact
external service the original method depended on had been discontinued
entirely, a case (`Linoleate`) where PubChem correctly resolves a name but
to a different, valid registration of the same compound than the one a
curated reference table happens to use, and - the largest single source of
disagreement with the published result - a systematic tendency for
automated name resolution to default to a different stereoisomer than a
human curator would pick from the same bare name (Section
\@ref(stereoisomers)). None of these are visible in a prose method
description; all of them are immediately visible, explainable, and at
least partially fixable once the workflow is code that actually runs.

# Learnings and Benefits of MetMashR

- No code was published; only the data was provided on Zenodo, despite the manuscript describing a specific multi-step pipeline (PubChem Identifier Exchange + CTS + manual curation). Using MetMashR, the pipeline itself would have been the artifact. A single, versioned workflow that can be shared, re-run, and cited alongside the paper, rather than a descriptive text that cannot be executed.
- The Chemical Translation Service (CTS) used in the paper is now fully shut down, and the CTS-lite replacement doesn't support the required name-to-id lookup. Using MetMashr, online lookups can be cached and version tracked, so that even when a service is no longer available previous results are still reproducible.
- The conjugate acid/base collapsing rule (using the first 14 characters of the InChIKey) is under-specified in the paper, making it difficult to reproduce exactly. Using MetMashr the code is inspectable and testable.
- Stereochemistry decisions were made by human curators with no visible trail. Several published names carry stereo descriptors the original reported data never had; they were added during "manual curation."  Using MetMashR, we incorporated a flag into the workflow to track where these discrepencies arise, so it's easier to see where manual curation may need to take place.
- In the paper, a mechanism to detect when two translation sources disagree is not provided. Their pipeline picks one answer (PubChem or CTS) without providing the selection logic or a record of the ambiguity. Using MetMashr, `prioritise_columns()` clearly captures which result get priority over the other and keeps a record of all results as well as the final resolved values.
- The case study in the paper is asserted to be reproducible but this is not demonstrated. A reviewer or reader has no way to check the result against the without rebuilding the entire pipeline from scratch using only the method text to guide them. Using MetMashr, the complete workflow can be run in a single step using data obtained from Zenodo and.or Metabolomics Workbench. Venn diagrams and other charts can be used to explore the resulting harmonised table and compare it to the table provided by the paper to explore and diagnose differences caused by e.g. using different id lookup services.

# References

Villalba H, Llambrich M, Gumà J, Brezmes J, Cumeras R. A Metabolites
Merging Strategy (MMS): Harmonization to Enable Studies' Intercomparison.
Metabolites 2023;13(12):1167. https://doi.org/10.3390/metabo13121167
