---
title: 'Case study: reproducing a Metabolite Merging Strategy with MetMashR'
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float: true
  word_document:
    toc: true
vignette: "%\\VignetteEncoding{UTF-8} %\\VignetteIndexEntry{Case study: reproducing
  a Metabolite Merging Strategy with MetMashR} %\\VignetteEngine{knitr::rmarkdown}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    fig.path = paste0(tempdir(), "/mms_case_study_figures/")
)

.DT <- function(x) {
    if (knitr::is_html_output()) {
        DT::datatable(
            x,
            options = list(scrollX = TRUE, pageLength = 10, dom = "ftip"),
            rownames = FALSE
        )
    } else {
        knitr::kable(x)
    }
}
```

```{r setup, message = FALSE}
library(struct)
library(MetMashR)
library(ggplot2)
```

# Introduction

Villalba et al. (2023) describe a three-step procedure for harmonising
metabolite annotations reported by different studies, databases and
publications into one consistent, cross-referenced dataset. The main steps
of their "Metabolite Merging Strategy" (MMS) are:

1. **Translation and merging** - convert identifiers from each source (e.g.
   a name, an InChIKey, or some other identifier) into an InChIKey, and
   merge records with the same key.
2. **Attribute retrieval** - for each InChIKey, retrieve descriptive
   attributes (such as name, molecular weight, molecular formula),
   identifiers for cross-references to other databases (e.g. PubChem CID,
   SMILES, ChEBI, HMDB, KEGG, LipidMaps, DrugBank, CAS) and chemical
   ontology (kingdom, superclass, class).
3. **Manual curation** - collapse conjugated acid/base forms of the same
   compound, retrieve missing attributes, and resolve duplicate/conflicting
   synonyms.

The worked example included in the original publication harmonises urinary
asthma metabolites reported across three Metabolomics Workbench (MWB)
studies (ST001039, ST001048, ST001317), the Human Metabolome Database, and
six literature articles.

This vignette reproduces the MWB portion of that case study as a single
MetMashR model sequence i.e. one workflow object built by chaining
MetMashR steps together, and checks the result against results tables
included in the supplementary information of the original article.

# Importing the source data

Each study's reported metabolite list is pulled directly from Metabolomics
Workbench with `mwb_study_source()` and the three tables are joined with
`combine_sources()`.

Only the `metabolite_name` column (the name as originally reported by each
study) is carried forward into the workflow. Metabolomics Workbench
does attach a `refmet_name` to most rows automatically at deposit time, and
occasionally a `pubchem_id`, but we exclude these for the purposes of
testing our MMS implementation.

```{r step0}
# the studies to import
study_ids <- c("ST001039", "ST001048", "ST001317")

# read each source
sources <- lapply(
    study_ids,
    function(id) read_source(mwb_study_source(source = id, tag = id))
)

# use MetMashR to merge all sources
combine_step <- combine_sources(
    source_list = list(),
    keep_cols = "metabolite_name",
    source_col = "source"
)

# apply the merge
combine_step <- model_apply(combine_step, sources)

# extract the merged table
combined <- predicted(combine_step)

# report number of records
nrow(combined$data)
```

# Building the model sequence {#step1}

The remaining steps are chained into a single model sequence and this
vignette runs all steps. To avoid overburdening the public APIs involved
every time that happens, every step that queries an external service is
given a preexisting `rds_cache` generated when the workflow was first run.
Rerunning the workflow uses these caches instead of re-querying. In
practice, you would typically omit `cache` (or point it at your own file)
so that your queries populate it as you go. Caching leaves a
persistent local record of what a service actually returned, which is
useful given that services used here can be discontinued or changed
without notice (as happened to CTS).

```{r cache-setup}
# Function that returns an rds_cache by name
cached <- function(name) {
    rds_cache(source = file.path(
        system.file("cached", package = "MetMashR"),
        paste0("mms_case_study_", name, ".rds")
    ))
}
```

## Normalising synonyms

Molecules have many synonyms and resolving them to the same identifier can
be very challenging. We apply a small number of normalisation rules, using
MetMashR "dictionaries", to maximise the chances of a matching query,
before attempting translation:

- remove record-keeping that isn't part of the
  chemical name (e.g. `"Dihydrobiopterin_R1"` becomes `"Dihydrobiopterin"`)
- remove short trailing batch-code markers (e.g. `"D,L Glucosamine (B)"` becomes
  `"D,L Glucosamine"`)
- where two alternative names are joined by a slash, keep the first 
  (e.g. `"MiBP (Mono-isobutyl phthalate)/MBP  (Mono-n-butyl phthalate)"` becomes 
  `"MiBP (Mono-isobutyl phthalate)"`)

```{r step0.51}
tidy_dictionary <- list(
    # trailing replicate/fragment label, e.g. "_R1", "_R2"
    list(pattern = "_[A-Za-z][0-9]+$", replace = ""),
    # remove short (<=3 character) trailing brackets batch/replicate codes
    list(pattern = "\\s\\([A-Za-z0-9]{1,3}\\)\\s*$", replace = ""),
    # keep only the first of two alternative names joined by a slash
    list(pattern = "\\)/.*$", replace = ")")
)
```
  
- extract a name when one is present alongside its abbreviation (e.g.
  `"MEP (Mono-ethyl phthalate)"` becomes `"Mono-ethyl phthalate"`)
  
```{r,step0.52}
abbrev_dictionary <- list(
    list(
        pattern = "\\s\\([^()]+\\)\\s*$",
        replace = function(inp) {
            m <- regmatches(
                inp$x,
                regexec("\\s\\(([^()]+)\\)\\s*$", inp$x)
            )
            if (length(m[[1]]) > 1) trimws(m[[1]][2]) else inp$x
        }
    )
)
```

- strip racemic/stereo markers. MetMashR's built-in `racemic_dictionary`
  removes all stereo and racemic markers (e.g. `"D/L "`, `"DL-"` and their
  mirror forms, `+/-`, etc).
  
- expand the `Lyso-` glycerophospholipid class prefix to LIPID MAPS
  shorthand (`"LysoPC"` becomes `"LPC"`).

```{r, step0.53}
# dictionary to convert lyso-lipids to LIPID MAPS shorthand
lyso_dictionary <- list(
    list(pattern = "^Lyso-?PC", replace = "LPC", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PE", replace = "LPE", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PI", replace = "LPI", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PG", replace = "LPG", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PS", replace = "LPS", perl = TRUE, ignore.case = TRUE),
    list(pattern = "^Lyso-?PA", replace = "LPA", perl = TRUE, ignore.case = TRUE)
)
```


The prepared dictories are used by `normalise_strings` to clean the text in 
specified `search_column` and return the result in 
the `output_column`.

```{r,step0.54}
# workflow step to apply the cleaning steps to the metabolite names
normalise_step <- normalise_strings(
    search_column = "metabolite_name",
    output_column = "name_normalised",
    dictionary = c(
        tidy_dictionary,
        abbrev_dictionary,
        racemic_dictionary, # included in MetMashR
        lyso_dictionary
    )
)
```

## Translation and merging (MMS Step 1)

Villalba et al. used the PubChem Identifier Exchange Service for names, and
the Chemical Translation Service (CTS) to translate other identifier types.
CTS has since been closed and its replacement, CTS-Lite, does not support
name-to-identifier translation. This vignette uses the Metabolomics
Workbench RefMet service instead.

`pubchem_id_exchange()` and `mwb_refmet_lookup()` are queried for a
match to every (normalised) name. `prioritise_columns()` then picks a
winner per row, preferring RefMet's result when one exists.

For clarity in the vignette, we will not build the complete sequence in one
go. Instead, we collect all steps in a list and then build the sequence via
`Reduce` at the end. Here, we include a step to indicate where the returned
InChIKey partially match and may therefore need manual review, a step to to 
exclude metabolites with no InChIKey from any source, and
another to merge metabolites with identical InChIKey into a single record.

```{r step1}
translate_step <- list(
    # query pubchem by synonym and return inchikey
    pubchem_id_exchange(
        query_column = "name_normalised",
        input_type = "synonyms",
        output_type = "inchikey",
        cache = cached("pubchem_id_exchange"),
        cache_mode = "offline"
    ),
    # query refmet by synonym
    mwb_refmet_lookup(
        query_column = "name_normalised",
        cache = cached("refmet"),
        cache_mode = "offline",
        delay = 0.5
    ),
    # flag where the first 14 characters of the inchikey match but the full
    # inchikey does not
    compute_column(
        input_columns = c("inchi_key_refmet", "inchikey_pubchem_id_exchange"),
        output_column = "stereo_ambiguous",
        fcn = function(x) {
            a <- x[[1]]
            b <- x[[2]]
            ambiguous <- substr(a, 1, 14) == substr(b, 1, 14) &
                substr(a, 1, 26) != substr(b, 1, 26)
            ambiguous[is.na(ambiguous)] <- FALSE
            ambiguous
        }
    ),
    # select refmet over pubchem where both exist
    prioritise_columns(
        column_names = c("inchi_key_refmet", "inchikey_pubchem_id_exchange"),
        output_name = "InChIKey",
        source_name = "InChIKey_source",
        source_tags = c("refmet", "pubchem")
    ),
    # exclude any results without an inchikey
    filter_na(column_name = "InChIKey", mode = "exclude"),
    # merge duplicate records
    combine_records(
        group_by = "InChIKey",
        default_fcn = fuse_unique(separator = " || "),
        fcns = list(
            metabolite_name = fuse_unique(separator = " || "),
            source = fuse_unique(separator = " || "),
            stereo_ambiguous = function(x) any(x, na.rm = TRUE)
        )
    )
)
```

`combine_records()` is a wrapper around `dplyr::reframe`. `fuse_unique` is a
helper function provided by MetMashR that in this case collapses multiple
synonyms (in the `metabolite_name` column) with the same InChIKey into a
single entry separated by a double pipe. For example, two entries for
`WQZGKKKJIJFFOK-GASJEMHNSA-N` might be combined into a single entry where
`metabolite_name` is `D-glucose || D-glucopyranose`. Thus records using different
synonyms for the same molecule are combined into a single record.

## Attribute retrieval (MMS Step 2)

Each translated InChIKey is enriched with additional information including:

- name, molecular weight/formula, InChI and SMILES from PubChem
- chemical ontology from ClassyFire, HMDB, ChEBI and LipidMaps identifiers
  from Metabolomics Workbench's compound database
- a KEGG identifier using `KEGGREST` (an R package that queries the KEGG api).

DrugBank ID and CAS number were included as part of the original paper but 
direct access the their api services require a private key, so they
are omitted here.

```{r step2}
attribute_step <- list(
    # query the inchikey with pubchem and return various properties
    pubchem_property_lookup(
        query_column = "InChIKey",
        search_by = "inchikey",
        property = c(
            "IUPACName", "MolecularWeight", "MolecularFormula",
            "InChI", "CanonicalSMILES", "Charge"
        ),
        cache = cached("pubchem_property"),
        cache_mode = "offline",
        delay = 0.4
    ),
    # query the SMILES (from pubchem) with classyfire and return class
    # information.
    classyfire_batch_lookup(
        query_column = "ConnectivitySMILES_pubchem",
        output_items = c("kingdom", "superclass", "class"),
        output_fields = "name",
        cache = cached("classyfire"),
        cache_mode = "offline",
        delay = 3
    ),
    # query the inchikey with MWB and return other cross reference
    # identifiers
    mwb_compound_lookup(
        input_item = "inchi_key",
        query_column = "InChIKey",
        output_item = c("hmdb_id", "chebi_id", "lm_id", "pubchem_cid"),
        cache = cached("mwb_compound"),
        cache_mode = "offline",
        delay = 0.5
    ),
    # query kegg with the pubchem CID and retrieve the kegg compound
    # identifier
    kegg_lookup(
        get = "compound",
        from = "pubchem",
        query_column = "CID_pubchem",
        cache = cached("kegg"),
        cache_mode = "offline"
    )
)
```

## Manual curation (MMS Step 3)

Villalba et al. describe collapsing conjugated acid/base forms using "the
first 14 characters" of the InChIKey (the connectivity/skeleton hash only).
We therefore group on the 14-character skeleton, treating
stereochemistry as unresolved. LC-MS generally cannot separate stereoisomers 
without a dedicated chiral
method, so a reported name's `D-`/`L-` prefix, where present, usually
reflects a biological assumption rather than an experimental finding.

```{r step3}
# create model sequence
curation_step <- list(
    # extract inchikey connectivity/skeleton only (first 14 characters) -
    # stereo-blind grouping, see prose above
    compute_column(
        input_columns = "InChIKey",
        output_column = "inchikey_skeleton",
        # compute_column passes fcn a 1-column data.frame, not a bare vector
        fcn = function(x) substr(x[[1]], 1, 14)
    ),
    # combine records with the same inchikey skeleton
    combine_records(
        group_by = "inchikey_skeleton",
        default_fcn = fuse_unique(separator = " || "),
        fcns = list(stereo_ambiguous = function(x) any(x, na.rm = TRUE))
    ),
    # remove any duplicate records
    unique_records()
)
```


## The complete sequence

Conceptually, the entire workflow described above is one continuous chain
of steps that could be joined using the plus "+" operator. In practice, due
to its length and for clarity in the vignette, it is assembled from the
four lists of models defined in the previous sections (`normalise_step`,
`translate_step`, `attribute_step`, `curation_step`) with `Reduce`.

```{r full-workflow}
# create model sequence
mms_workflow <- Reduce(
    `+`,
    c(list(normalise_step), translate_step, attribute_step, curation_step)
)

# apply model sequence to merged sources
mms_workflow <- model_apply(mms_workflow, combined)

# extract result
result <- predicted(mms_workflow)

# report number of records
nrow(result$data)
```

# What the harmonised table looks like

`result$data` is a single flat table with one row per unique compound
including every attribute from Steps 1-3 as a column. The InChIKey and its
provenance, PubChem/RefMet identifiers, ClassyFire ontology, and
cross-references from MWB, ChEBI, HMDB, LipidMaps and KEGG are all
included. A cleaned-up subset of the columns is shown below (the full table
also carries e.g. the raw ClassyFire class descriptions, InChI strings and
SMILES, omitted here for readability) for the first 10 records:

```{r result-table}
preview_cols <- c(
    "metabolite_name", "InChIKey", "IUPACName_pubchem",
    "MolecularFormula_pubchem", "MolecularWeight_pubchem",
    "hmdb_id_mwb", "chebi_id_mwb", "pubchem_cid_mwb", "compound_kegg", "lm_id_mwb",
    "kingdom.name_cfb", "superclass.name_cfb", "class.name_cfb",
    "stereo_ambiguous", "source"
)
.DT(head(result$data[, preview_cols], 10))
```

# Comparison against the published result

The final table of the original publication is published as supplementary
material on Zenodo (10.5281/zenodo.8226097). It's imported here with
`BiocFileCache_database()` and restricted to compounds attributed to the
three MWB studies (the origial publications HMDB and literature sources are out of scope
for this vignette).

The published InChIKeys are still stereo-specific, so the
comparison below is made at the same 14-character skeleton level used for
grouping in Step 3 of the vignette. Here we implement a small MetMashR
workflow to process the imported data.

```{r comparison, out.width = "70%", fig.width = 8, fig.height = 6}
# url to the content
zenodo_url <- paste0(
    "https://zenodo.org/api/records/8226097/files/",
    "MDM_Suppl_vSubmitted.xlsx/content"
)

# prepare BiocFileCache object
gold_db <- BiocFileCache_database(
    source = zenodo_url,
    resource_name = "MMS_zenodo_suppl",
    bfc_fun = cache_as_is,
    import_fun = function(path) {
        # the Zenodo API URL has no .xlsx extension; openxlsx needs one
        tmp <- tempfile(fileext = ".xlsx")
        file.copy(path, tmp, overwrite = TRUE)
        on.exit(unlink(tmp))
        openxlsx::read.xlsx(tmp, sheet = "Table S3", colNames = TRUE)
    }
)

# import as an annotation_source
gold <- read_source(gold_db)

# restrict to the three MWB studies, then derive the same
# inchikey skeleton used in Step 3
gold_workflow <- 
    filter_labels(
        column_name = "Reference", 
        labels = study_ids, mode = "include") + 
    compute_column(
        input_columns = "InChIKey", 
        output_column = "inchikey_skeleton",
        fcn = function(x) substr(x[[1]], 1, 14))

# apply the model the imported source
gold <- predicted(model_apply(gold_workflow, gold))

# set labels for that get used for by the venn chart
gold$name <- "MMS (Villalba et al.)"
gold$tag <- "MMS (Villalba et al.)"

# The vignette result is already an annotation_source with an inchikey_skeleton
# column from Step 3 so it can be used directly
result$name <- "MetMashR"
result$tag <- "MetMashR"

# compare the two inchikey skeleton columns using a venn diagram
C <- 
    annotation_venn_chart(
        factor_name = "inchikey_skeleton", 
        labels = TRUE, 
        legend = TRUE,
        fill_colour = ".group"
)
chart_plot(C, result, gold)
```

# Characterising the mismatch {#stereoisomers}

```{r mismatch-check, echo = FALSE}
our_keys <- unique(result$data$inchikey_skeleton)
gold_keys <- unique(gold$data$inchikey_skeleton)

only_ours <- setdiff(our_keys, gold_keys)
only_gold <- setdiff(gold_keys, our_keys)
n_shared <- length(intersect(our_keys, gold_keys))
pct_ours <- sprintf("%.1f", 100 * n_shared / length(our_keys))
pct_gold <- sprintf("%.1f", 100 * n_shared / length(gold_keys))
```

At the skeleton level, `r n_shared` compounds (`r pct_ours`% of MetMashR's
`r length(our_keys)`, `r pct_gold`% of the paper's `r length(gold_keys)`)
are shared between the two results, leaving `r length(only_ours)` only in
MetMashR's result and `r length(only_gold)` only in the original publication. This is a high
level of agreement given every compound was resolved from a synonym.

The remaining differences are due to a few names that  don't
resolve to anything in PubChem or RefMet's synonym databases even once
cleaned.

# Learnings and Benefits of MetMashR

- No code was included in the original paper; only the final result was provided on Zenodo. Using MetMashR, we have reproduced the MMS workflow. It is now a single, reproducible workflow that can be shared, re-run, and cited, rather than descriptive text that cannot be executed.
- The Chemical Translation Service (CTS) used in the paper is now fully shut down, and the CTS-lite replacement doesn't support the required name-to-id lookup used by the original publication. Using MetMashr, online lookups can be cached and version tracked, so that even when a service is no longer available previous results are still reproducible.
- The conjugate acid/base collapsing rule (using the first 14 characters of the InChIKey) is under-specified in the paper, making it difficult to reproduce exactly. Using MetMashr the code is inspectable and testable, allowing the method to be inspected even when not fully described in text.
- Stereochemistry decisions were made in the original publication by human curators with no visible trail. Several published names carry stereo descriptors the original reported data never had; they were added during "manual curation." Using MetMashR, we incorporated a flag into the workflow to track where these discrepencies arise, so it's easier to see where manual curation may need to take place.
- In the paper, a mechanism to detect when two translation sources disagree is not provided. Their pipeline picks one answer (PubChem or CTS) without providing the selection logic or a record of the ambiguity. Using MetMashr, `prioritise_columns()` clearly captures which result get priority over the other and keeps a record of all results as well as the final resolved values.
- The case study in the paper is asserted to be reproducible but this is not demonstrated. A reviewer or reader has no way to check the result against the without rebuilding the entire pipeline from scratch using only the method text to guide them. Using MetMashr, the complete workflow can be run in a single step using data obtained from Zenodo and/or Metabolomics Workbench. Venn diagrams and other charts can be used to explore the resulting harmonised table and compare it to the table provided by the paper to explore and diagnose differences caused by e.g. using different id lookup services.

# References {-}

Villalba H, Llambrich M, Gumà J, Brezmes J, Cumeras R. A Metabolites
Merging Strategy (MMS): Harmonization to Enable Studies' Intercomparison.
Metabolites 2023;13(12):1167. https://doi.org/10.3390/metabo13121167

# Session Info

```{r}
sessionInfo()
```
